---
title: "Toronto’s Homelessness Crisis"
subtitle: "A Preliminary Analysis of Deaths by Month"
author: 
  - Marcus Emmanuel Barnes
thanks: "Code and data are available at: [https://github.com/MarcusBarnes/Toronto_Homeless_Deaths](https://github.com/MarcusBarnes/Toronto_Homeless_Deaths)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)

# Read in the data.
data <- read_csv(here("data/01-raw_data/raw_homeless_deaths_by_month_data.csv"))

head(data)
```


# Introduction
There is a general sense that the homeless crisis in Toronto is getting worse [@citeCTVNewsToronto].  A recent study [@citeRichard2024disparities] indicates that, on average, homeless Torontonians are 17 years younger than those who have not experienced homelessness.   As the lead author Lucie Richard lements in an interview about the study, there is a lack of cause of death information to help better pinpoint interventions to reduce premature death [@citeCTVNewsToronto]. Given the lack of cause of death information, it is worthwhile exploring the data that does exist.  Toronto Public Health (TPL) has been collecting data on homeless deaths by month since January 2017.  We will explore this data set and determine what stories the data tells us. These stories may help inform leadership of governmental and non-govenmental agencies better utilize their scarce resource in the creation of more effective intervention initiatives reduce deaths among Toronto's homeless population.

# Data {#sec-data}
In this report, we explore the 'Deaths of People Experiencing Homelessness' data set [@citeTorontoHomelessDeaths] which was collected using the Open Data Toronto Library [@citeOpenDataToronto] utilizing the statistiacal programming software R [@citeR].

## Deaths of People Experiencing Homelessness Statistics
For the purposes of this dataset, Toronto Public Health (TPH), defines homelessness as "the situation of an individual or family without stable, permanent, appropriate housing, or the immediate prospect, means and ability of acquiring it" [@citeTorontoHomelessDeaths]. Since January 2017, TPH has counted the number of deaths of people experiencing homelessness by year and month. The deaths are repoted to TPH by the Shelter, Support and Housing Administration (SSH), community partner organizations and the Coroner's Office [@citeTorontoHomelessDeaths].

# Analysis

## Model set-up



### Model justification



# Results
```{r}
#| message: false
#| echo: false
#| label: summary_statistics
#| fig-cap: Summary statistics

summary(data$Count)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check


## Diagnostics




\newpage


# References


