---
title: "Toronto’s Homelessness Crisis"
subtitle: "A Preliminary Analysis of Deaths by Month"
author: 
  - Marcus Emmanuel Barnes
thanks: "Code and data are available at: [https://github.com/MarcusBarnes/Toronto_Homeless_Deaths](https://github.com/MarcusBarnes/Toronto_Homeless_Deaths)."
date: today
date-format: long
abstract: "There is a general sense that the Toronto Homeless situation is getting worse.  Given the lack of cause of death information to help better pinpoint interventions to reduce premature death of Homeless Torontonians, we carry out a preliminary exploration of a dataset we do have: the 'Deaths of People Experiencing Homelessness' dataset published by Toronto Public Health available via Open Data Toronto. The data shows that COVID-19 lockdowns increased homeless deaths during that period. Using the data, we create a forecast of 2024 homeless deaths using the Prophet procedure.  We close with a discussion of the limitations of our preliminary analysis and forecase and outline next steps."
format: pdf
number-sections: true
bibliography: references.bib

---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(ggplot2)
library(dplyr)
library(lubridate)
library(prophet)

# Read in the data.
data <- read_csv(here("data/01-raw_data/raw_homeless_deaths_by_month_data.csv"))

head(data)
```


# Introduction
There is a general sense that the homeless crisis in Toronto is getting worse [@citeCTVNewsToronto]. A recent study [@citeRichard2024disparities] indicates that, on average, homeless Torontonians are 17 years younger than those who have not experienced homelessness. As the lead author Lucie Richard lements in an interview about the study, there is a lack of cause of death information to help better pinpoint interventions to reduce premature death [@citeCTVNewsToronto]. Given the lack of cause of death information, it is worthwhile exploring the data that does exist.  Toronto Public Health (TPL) has been collecting data on homeless deaths by month since January 2017. We explore this data set and determine what stories the data tells us. These stories may help inform leadership of governmental and non-governmental agencies better utilize their scarce resource in the creation of more effective intervention initiatives reduce deaths among Toronto's homeless population.

# Data {#sec-data}
In this report, we explore the 'Deaths of People Experiencing Homelessness' data set [@citeTorontoHomelessDeaths] which was collected using the Open Data Toronto Library [@citeOpenDataToronto] utilizing the statistical programming software R [@citeR].

## Deaths of People Experiencing Homelessness Statistics
For the purposes of this dataset, Toronto Public Health (TPH), defines homelessness as "the situation of an individual or family without stable, permanent, appropriate housing, or the immediate prospect, means and ability of acquiring it" [@citeTorontoHomelessDeaths]. Since January 2017, TPH has counted the number of deaths of people experiencing homelessness by year and month. The deaths are reported to TPH by the Shelter, Support and Housing Administration (SSH), community partner organizations and the Coroner's Office [@citeTorontoHomelessDeaths].

# Analysis

## Summary Statistics
In the existing data, the monthly death counts range from 1 to 26.  The median monthly death count is 11 and the mean death count is approximately 12.7.  The standard deviation is 5.06.
```{r}
#| message: false
#| echo: false
#| label: summary_statistics
#| fig-cap: Summary statistics

#summary(data$Count)
#mean_count = mean(data$Count, na.rm = TRUE)
#median_count = median(data$Count, na.rm = TRUE)
#sd_count = sd(data$Count, na.rm = TRUE)



```
## Distribution of Monthly Death Counts
In @fig-distribution-of-monthly-deaths, we see the distribution of monthly death counts.
```{r}
#| message: false
#| echo: false
#| label: fig-distribution-of-monthly-deaths
#| fig-cap: Distribution of Monthly Deaths

# Convert Month.of.death to a date format
data$Date <- as.Date(paste(data$Year.of.death, match(data$Month.of.death, month.name), "01", sep = "-"))

# Histogram of 'Count'
ggplot(data, aes(x = Count)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Distribution of Monthly Death Counts", x = "Count", y = "Frequency")
```

##  
```{r}
#| message: false
#| echo: false
#| label: fig-death-count-over-time
#| fig-cap: Death Count over Time.

# Convert Month.of.death to a date format
data$Date <- as.Date(paste(data$Year.of.death, match(data$Month.of.death, month.name), "01", sep = "-"))

# Time series plot
ggplot(data, aes(x = Date, y = Count)) +
  geom_line() +
  labs(title = "Death Count Over Time", x = "Date", y = "Count")
```

## Forecast Model set-up
Using the Prophet package[@citeProphetPackage], we can attempt to forecast what will happen over the next 12 months.

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-prophet-forecast
#| fig-cap: Forecast of the next 12 months using Prophet forecast procedure.

# Convert Year and Month into a datetime format
# ToDo - you're repeating this.  Process the data before including in
# the main paper Quarto file.
data$Date <- ymd(paste(data$Year.of.death, data$Month.of.death, "01"))

# Arrange the data by Date
data <- data %>% arrange(Date)

# Prepare the data for Prophet
prophet_data <- data %>%
  select(Date, Count) %>%
  rename(ds = Date, y = Count)

# Fit the Prophet model
model <- prophet(prophet_data)

# Make a future dataframe for the next 12 months
future <- make_future_dataframe(model, periods = 12, freq = "month")

# Forecast the future values
forecast <- predict(model, future)

# Plot the forecast
plot(model, forecast)

# Extract the relevant columns for the next 12 months
forecast_table <- forecast %>%
  select(ds, yhat) %>%
  filter(ds > max(prophet_data$ds))

# Print the forecast table
#print(forecast_table)
```
In particular, we have the following monthly forecast for 2024 (where we have rounded the predicted number up to a whole number):
```{=latex}

\begin{table}[h!]
\centering
\begin{tabular}{|c|c|}
\hline
\textbf{Month} & \textbf{Forecasted Deaths} \\
\hline
January & 21 \\
February & 16 \\
March & 14 \\
April & 15 \\
May & 22 \\
June & 15 \\
July & 19 \\
August & 20 \\
September & 15 \\
October & 16 \\
November & 20 \\
December & 20 \\
\hline
\end{tabular}
\caption{Forecasted death counts of homeless Torontonians for 2024 by month.}
\label{table:forecast}
\end{table}

```
### Model justification
According to The Prophet package CRAN page, the forecasting procedure is "Prophet is robust to missing data and shifts in the trend, and typically handles outliers well" [@citeProphetPackage].  


# Forecasting Results
Using the Prophet forecasting procedure have the following monthly forecast for 2024 (where we have rounded the predicted number up to a whole number):
```{=latex}

\begin{table}[h!]
\centering
\begin{tabular}{|c|c|}
\hline
\textbf{Month} & \textbf{Forecasted Deaths} \\
\hline
January & 21 \\
February & 16 \\
March & 14 \\
April & 15 \\
May & 22 \\
June & 15 \\
July & 19 \\
August & 20 \\
September & 15 \\
October & 16 \\
November & 20 \\
December & 20 \\
\hline
\end{tabular}
\caption{Forecasted death counts of homeless Torontonians for 2024 by month.}
\label{table:forecast}
\end{table}

```
# Discussion

## Impact of COVID-19 lockdowns.

From the plot of Count over Time (@fig-death-count-over-time), there is a clear increase in homeless dealths during that period. This can be attributed to reduces services for and increased social isolation of homeless Torontonians during this period. <!-- cite relevant sources to substantiate. -->

## Forecasts for Toronto Funeral Expense Budget Allocation {#sec-first-point}
Our forecast of monthly deaths for 2024 does not help us in designing direct interventions to avoid these deaths, but it can assist the city of Toronto allocate appropriate funds for the funeral expenses of the deceased [@citeCityofToronto_HealtSupport].

## Weaknesses

### Lack of Data
This data set only goes back to 2017.  As additional data is collected each year, certainty on discernable trends will increase.

### Forecast Models
Other time series forecasting models such as ARIMA or Holt-Winters available in the forecast package could be considered [@citeForecastPackage].  Each has its pros and cons. Prophet was primarily selected for its ease of use.

## Next Steps

## Enriching and Complementing the Data
A next step is to look at how we can enrich the relatively basic dataset under study with additional data.  This enrichment may provide more insight into the collected rates of deaths amongs homeless Torontonians and may provide information which can help form improved interventions.

### Utilizing Time Series Data
Based on the time series forecast presented in this report, follow up next year to determine the accuracy of the forecasted trend.  Were other forecasting techniques such as ARIMA or Holt-Winters ultimately more correct?  What can we now say with an additional year of collected data?

The concept of motifs in time series data may provide additional insight into less evident patterns in the data.  There is a powerful Python library created for exploring motifs in time series data called STUMPY [@citeLaw2019stumpy].  By using the `reticulate` package [@citeReticulatePackage], it's possible to use STUMPY within R as demonstrated in a relevant tutorial [@citeStumpr].

<!--
# Saving for future iterations of enhacements.
\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check


## Diagnostics
-->



\newpage


# References


